---
title: "STAT40830 Advanced Data Programming with R - Homework 1"
author: "Aoife Doyle Student No: 15419052"
format: 
  pdf:
    geometry: 
      - top=15mm
      - bottom=15mm
      - left=12mm
      - right=12mm
      - heightrounded
    fontsize: 10pt  
    linestretch: 1.1  
execute:
  echo: true  
  eval: true  
---


## Dataset Description

For this assignment we have chosen to use the built in **Boston Housing** 
dataset from the MASS package which contains housing data for various suburbs 
of Boston. Utliising the `?` and `str()` functions we are able to call the R 
help file and looking at the structure of our dataset we are able to determine 
that our dataset has 14 variables and 506 observations. It should be noted that 
almost all of our variables are numeric in nature aside from our chas 
(river dummy variable) and rad (an index variable for access to urban highways) 
which are both classified as being of type interger. 

```{r}
library(MASS) # loading in the MASS package
?Boston # calls the R help file to see a brief descriptions of the datasets 
# variables
str(Boston) # function that allows us to see the structure of our dataset and 
# its variables
head(Boston, 10) # prints the first 10 rows of our dataset

```

## Descriptive Statistics Analysis

Examing the summary statistics below reveals a wide range of values across key 
housing attributes for our Boston dataset. For instance, examining our variable 
for the crime rate per capita (crim) shows that crime rates in Boston can vary
drastically from as low as 0.006 to nearly 89, indicating stark contrasts in 
different neighborhoods safety levels. The median value of homes (medv) ranges 
from **$5,000** to **$50,000**, with a mean of **$22,530**, suggesting some 
skewness and possible value capping at the upper end of our distributions tails.

There is also variation in socio-economic indicators such as the percentage of 
lower-status residents (lstat), which ranges from *1.73%* to *37.97%*, and 
property tax rates (tax), which range from *187* to *711*. Additionally, the 
average number of rooms per dwelling (rm) is around *6.28*, with most homes 
having between *5* and just under *7* rooms. These variables provide valuable 
insights into the structural and environmental characteristics influencing 
housing prices in the Boston area.

```{r}
summary(Boston) # prints summary statistics for each variable in our dataset
```

In line with our summary statistics @fig-boxplot below presents boxplots 
of all variables in our Boston dataset. For several variables such as crime rate
(crim), property tax (tax), and the percentage of lower-income residents (lstat), 
we find that they exhibit strong right-skewed distributions with significant 
outliers which suggests the presence of extreme values in certain Boston 
neighborhoods. In contrast, variables such as number of rooms (rm) and the 
pupil-teacher ratio (ptratio) appear to be more symmetrically distributed. 

Overall we find that our plot also highlights the wide variability across 
features, which is line with previous findings around substantial variation in 
socio-economic and environmental disparities among different neighbourhoods. 
However,to further understand how these variables are interrelated, we now 
turn to the correlation matrix in @fig-corr, which provides deeper 
insights into the strength and direction of linear relationships between our 
datasets key variables.

```{r}
#| label: fig-boxplot
#| fig-cap: "Boxplot of our Boston Housing datasets Variables"
#| fig-width: 7
#| fig-height: 4

library(ggplot2)  # Functions used for visualization
library(reshape2) # Functions used for reshaping our data

# Utilising the melt function to reshape the data into a long format for data 
# visualisation
boston_long <- melt(Boston)

# Creating a boxplots for our variables
ggplot(boston_long, aes(x = variable, y = value)) +
  geom_boxplot(fill = "lightblue", color = "darkblue", outlier.color = "darkred") +
  theme_minimal() +
  labs(title = "Figure 1: Boxplot of our Boston Housing Variables",
       x = "Variable", y = "Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

@fig-corr below presents the correlation matrix for our Boston Housing 
dataset in order to highlight the strength and direction of relationships 
between our numeric variables. Overall we find that the strongest positive 
correlation is between rad (index of accessibility to radial highways) and tax 
(property tax rate), with a coefficient of *+0.91*, suggesting that areas with 
greater access to highways are significantly more likely to have higher property 
taxes.

In contrast, we observed that our strongest negative correlation was between 
lstat (percentage of lower-status population) and medv (median home value), with
a correlation coefficient of *–0.74*, thus reinforcing the fact that areas with 
higher concentration of economically disadvantaged residents will tend to have 
lower housing prices.

Lastly we find that houses in proximity to the Charles River (chas) shows the 
weakest correlations among all of our variables, particularly with zn (the 
proportion of land zoned for lots over 25,000 sq.ft.) with *-0.04*, indicating 
that being close to a river has little to no linear relationship with zoning and 
other housing characteristics in our dataset overall.

In sum our correlation analysis highlights strong impact that key socioeconomic 
and structural factors such as access to highways, number of rooms, and income 
level can have on influencing housing values in Boston, while other variables 
like proximity to the river appear to have minimal to no impact.


```{r}
#| label: fig-corr
#| fig-cap: "Correlation Matrix for our Boston Housing Dataset"
#| fig-width: 7
#| fig-height: 4

# Creating our correlation matrix for our heat map
numeric_boston <- Boston[, sapply(Boston, is.numeric)]
corr_matrix <- cor(numeric_boston)

library(ggcorrplot) # functions for reordering the correlation matrix and 
# displays the significance level on the plot.

# Creating a Heatmap using ggcorrplot to see the relationship between variables
ggcorrplot(corr_matrix, 
           method = "square", # squares instead of circles for data visuals
           type = "lower", # show only lower triangle
           lab = TRUE, # add in correlation variables
           lab_size = 3, 
           colors = c("darkred", "white", "darkblue"),
           title = 
             "Figure 2: Correlation Matrix for our Boston Dataset") +
  theme_minimal()


```